A Linear Technique to Predict Non-Normal Turbulence Applied to a Magnetized Plasma

### 2013-12-19 Transient_growth_prl.tex
Dec 19

\documentclass[showpacs,preprintnumbers,amsmath,amssymb,superscriptaddress,aip]{revtex4-1}
\usepackage{graphicx}
%\setlength{\parindent}{0in}

@@

\title{A Linear Technique to Predict Non-Normal LAPD Turbulence}

@@

\begin{abstract}
In nonlinear dynamical systems with highly nonorthogonal linear eigenvectors, linear non-modal analysis is more useful than normal mode analysis in predicting turbulent properties. 
However, the non-trivial time evolution of non-modal structures makes quantitative prediction difficult. Here, we present a technique to overcome this difficulty, which includes
modelling the effect that the advective nonlinearities have on spatial turbulent structures. The nonlinearities are taken as a periodic randomizing force with
period consistent with critical balance arguments. We apply this technique to a model of drift wave turbulence in the Large Plasma Device (LAPD) 
[W. Gekelman \emph{et al.}, Rev. Sci. Inst. {\bf 62}, 2875 (1991)], where non-modal effects dominate the turbulent system.
We compare the resulting growth rate spectrum to that obtained from a nonlinear simulation, showing good quantitative agreement.
\end{abstract}

@@

\section{Introduction}

Normal mode analysis -- the calculation of eigenvalues and eigenvectors of a linearized dynamical system -- has been used for the last two centuries to solve a number of problems ranging from
heat conduction along a solid bar to quantum mechanical energy states of atoms. Despite its wide-ranging success, normal mode analysis has seemingly failed in particular instances,
particularly in predicting the onset of turbulence in many hydrodynamic flows, where the turbulence is called subcritical. 
The reason for this failure was not clearly explained until the early 1990's when Trefethen and others attributed the pitfalls of normal mode analysis to the non-normality of linear operators of
dynamical systems~\cite{trefethen1993,trefethen2005}. A non-normal operator or matrix is one that does not commute with its adjoint, and it has 
eigenvectors that are nonorthogonal to one another. One consequence of eigenvector nonorthogonality is that even when all eigenvectors are stable and decay exponentially under linear evolution, 
superpositions of eigenvectors can grow, albeit transiently.
A paradigmatic illustration of this transient growth process for two nonorthogonal eigenvectors may be seen if Figure 2 of a review paper by Schmid~\cite{schmid2007}.
Physically, this means that certain types of fluctuations of the laminar state can access free energy from background gradients even though normal mode fluctuations cannot.
When combined with nonlinear effects, this allows for sustained subcritical turbulence.
Such behavior is obscured by traditional normal mode analysis, which only effectively describes the long time asymptotic behavior of fluctuations under the 
action of the linear operator. Transient growth events, which can dominate nonlinear turbulent evolution, require initial-value (non-modal) calculations rather than normal mode calculations.

Non-modal analysis has been embraced by the hydrodynamics community over the past two decades in the attempt to explain and predict subcritical turbulence. But the plasma community
still heavily relies on normal mode analysis to inform turbulent predictions and observations, with a few notable exceptions~\cite{camargo1998,camporeale2010,schekochihin2012}. 
Furthermore, non-modal treatments thus far have generally been explanatory rather than predictive and have centered around the transition to turbulence in subcritical systems rather 
than on properties of fully-developed turbulence.
This paper takes up the task of developing a quantitative approach to predicting turbulent properties using only non-modal linear calculations. 
The basic approach we take is to solve the linear initial value problem using an ensemble of random initial conditions, 
and then calculate the average growth rate of the solutions up to one nonlinear decorrelation time. We take this nonlinear decorrelation time to be a characteristic \emph{linear} time.
This procedure produces an effective growth rate spectrum that can be used to predict turbulent properties such as saturation levels.
While the concepts behind this technique are general enough to be applied to many nonlinear dynamical systems, the details will vary from system to system, 
so we restrict our treatment to one particular turbulence model.

The model is designed to describe a drift wave turbulence experiment conducted in the Large Plasma Device (LAPD)~\cite{gekelman1991}. LAPD is a linear machine with a straight magnetic $\mathbf{B}$ field.
Due to its large dimensions and high collisionality, LAPD is apt for fluid modelling. Thus, we use a reduced Braginskii~\cite{Braginskii1965} 2-fluid model:

\beqar
\label{ni_eq}
\pdt N = - {\mathbf v_E} \cdot \grad N_0 - N_0 \gradpar \vpe + \mu_N \gradperp^2 N + S_N + \{\phi,N\}, \\
\label{ve_eq}
\pdt \vpe = - \fmie \frac{T_{e0}}{N_0} \gradpar N - 1.71 \fmie \gradpar T_e  + \fmie \gradpar \phi - \nue \vpe + \{\phi,\vpe \}, \\
\label{rho_eq}
\pdt \varpi = - N_0 \gradpar \vpe  - \nuin \varpi + \mu_\phi \gradperp^2 \varpi + \{\phi,\varpi \}, \\
\label{te_eq}
\pdt T_e = - {\mathbf v_E} \cdot \grad T_{e0} - 1.71 \frac{2}{3} T_{e0} \gradpar \vpe + \frac{2}{3 N_0} \kpe \gradpar^2 T_e  \nonumber \\
- \frac{2 m_e}{m_i} \nue T_e  + \mu_T \gradperp^2 T_e +  S_T + \{\phi,T_e\},
\eeqar
where $N_0$ and $N$ are the equilibrium and fluctuating density, $\vpe$ is the fluctuating parallel electron velocity, $\varpi \equiv \gradperp \cdot (N_0 \gradperp \phi)$ is the potential vorticity
of the fluctuating potential $\phi$, and $T_{e0}$ and $T_e$ are the equilibrium and fluctuating electron temperature. The equations employ Bohm normalizations: lengths are
normalized to the ion sound gyroradius $\rho_s$, times to the ion cyclotron time $\omega_{ci}^{-1}$, velocities to the sound speed $c_s$, densities to the equilibrium peak density, and electron
temperatures and potentials to the equilibrium peak electron temperature. The profiles $N_0$ and $T_{e0}$ and other parameters are all taken from experimental measurements. $\phi_0$ 
does not appear in the equations because we model one particular experiment in which the equilibrium radial electric field was nulled by boundary biasing~\cite{schaffner2012}.

The equations are global and partially linearized. The only nonlinearities retained are the advective nonlinearities, which are written with Poisson brackets. 
We add artificial diffusion and viscosity terms with small numerical coefficients ($\mu_N, \mu_\phi, \mu_T = 10^{-3}$)
for numerical stability in nonlinear simulations, which are performed with the BOUT++ code~\cite{dudson2009}. We use periodic axial and zero value radial 
boundary conditions. Further details of the model, including validation studies, may be found in the references~\cite{Popovich2010a,Popovich2010b,Umansky2011,friedman2012b,friedman2013},
though we mention here that the model reproduces all statistical properties of the experimental turbulence within a factor of two, and in most cases, much better than that.

The nonlinear simulation reveals a fascinating property of the turbulence -- it is dominated by a nonlinear instability process despite being linearly unstable to drift waves
~\cite{friedman2012b,friedman2013}.
This nonlinear instability process, which was discovered by Drake et al.~\cite{drake1995} in a similar plasma model to ours, works as follows: 
long ($k_\para=0$) convective filaments transport density across the equilibrium density profile, setting up $k_\para=0$ density filaments (the same applies to the temperature). 
These filaments are unstable to drift waves causing a secondary drift wave instability to grow on the filaments. 
These drift waves, which have finite $k_\para$ and propagate primarily radially, nonlinearly couple to one another and reinforce the convective filaments.

@@

\beq
\label{dEdt_def}
\diff{E(m,n,t)}{t} = Q(m,n,t) + D(m,n,t) + \sum_{m',n'} T(m,m',n,n',t).
\eeq
In this equation, $m$ and $n$ represent the azimuthal and axial Fourier mode numbers, respectively. 
Since Eqs.~\ref{ni_eq}-\ref{te_eq} are not explicitly dependent on $\theta$ and $z$, the energy equation is separable for every $m$ and $n$. 
Futher, $Q$ represents the linear energy exchange with the equilibrium (convective transport), $D$ the dissipation from irreversible processes
like collisions, and $T$ the nonlinear triad wave transfer. Since $\int f \{g,f\} dV = 0$, or equivalently $\sum_{m,m',n,n'} T(m,m',n,n',t)=0$, 
the Poisson bracketed advective nonlinearities in the model are energetically conservative and therefore do no inject or dissipate energy on the whole.
Consequently, the fluctuation energy originates completely from $Q$, the linear convective transport mechanism.

\begin{figure}
\centerline{\includegraphics[]{transient_n0_n1_growth}}
\caption{The linear evolution of the energy of several Fourier modes starting from a turbulent initial state. The $n=0$ curves have an initial period of transient growth before exponentially decaying,
while the $n=1,m=20$ curve transiently decays before growing exponentially.}
\label{transient_n0_n1_growth}
\end{figure}

This convective transport of density filaments is akin to the paradigmatic ``lift-up'' mechanism in hydrodynamic shear flows whereby streamwise vortices drive streamwise streaks~\cite{krommes1999}.
Both are transient growth processes. We see this in our simulations by following the evolution of the energy of several Fourier $m,n$ modes after turning off the nonlinearities in an already
turbulent simulation. A few representative $m,n$ modes are shown in Fig.~\ref{transient_n0_n1_growth}. The transient growth of the filamentary structures is seen in the $n=0$ modes, 
which grow transiently before decaying exponentially at the rate of their least stable eigenmode. This kind of behavior is rather unituitive to some because all $n=0$ linear eigenmodes are stable, 
and the normal mode paradigm has conditioned us to believe that a linear superposition of such eigenmodes should decay. But as the hydrodynamics community discovered, linear
superpositions of nonorthogonal vectors may behave unintuitively.
Notice now that the $n=1,m=20$ mode decays transiently before growing exponentially with growth rate of the most unstable eigenmode at $n=1,m=20$. This isn't quite as interesting as the $n=0$
behavior, and it can occur in normal systems. Futhermore, it has no parallel in truly subcritical turbulence, but it can be helpful for interpreting later results.

@@

\begin{figure}
\centerline{\includegraphics[]{spec_nl_gamma}}
\caption{The linear and turbulent growth rate $m$ spectra for $n=0$ (solid lines) and $n=1$ (dashed lines) Fourier components. The linear growth rates are those for the least stable eigenmodes at each
$n$ and $m$, while the turbulent growth rates represent $\pdiff{E}{t}|_{\rm{lin}}/2 E$ from the nonlinear simulation. The shaded region marks the $1 \sigma$ spread in the turbulent growth rate spectrum,
obtained from the distribution of growth rates calculated from the nonlinear simulation over a long time range.}
\label{spec_nl_gamma}
\end{figure}

To obtain the turbulent growth rate spectrum from the nonlinear simulation, we begin with the expression $\diff{E(m,n,t)}{t}\big|_{\rm{lin}} = Q(m,n,t) + D(m,n,t)$, 
which accounts for the total net energy injection at each $m,n$ into the fluctuations. 
An effective instantaneous growth rate $\gamma_{\rm{eff}}(m,n,t) = \diff{E(m,n,t)}{t}\big|_{\rm{lin}}/ 2 E(m,n,t)$ can then be defined. 
This effective growth rate is calculated from the spatial structure of the fluctuations, so it is well-defined for both turbulent fluctuations and eigenmode fluctuations. If calculated
from a linear simulation run for a sufficiently long time, $\gamma_{\rm{eff}}(m,n)$ is the same as the eigenmode growth rate of the least stable linear eigenmode for each $m,n$.
For turbulence, $\gamma_{\rm{eff}}(m,n)$ is generally not equal to an eigenmode growth rate, especially in the case where the system is non-normal.
We plot this growth rate in Fig.~\ref{spec_nl_gamma} for $n=0$ and $n=1$ components. The turbulent growth rate curves are calculated from this formula, 
averaged over a time of about $3 \times 10^{4} \ \omega_{ci}^{-1}$ during the saturated turbulent phase of the nonlinear simulation: $\gamma_{\rm{Turb}}(m,n) = \int \gamma_{\rm{eff}}(m,n,t) dt$. 
We show this turbulent growth rate along with the $1 \sigma$ contour, which indicates how greatly $\gamma_{\rm{eff}}(m,n,t)$ varies over time.
Along with these, the growth rates of the least stable (or most unstable) linear eigenmodes are shown.

As already stated, one sees from Fig.~\ref{spec_nl_gamma} that the linear eigenmodes are unstable for $n=1$ and stable for $n=0$ as expected for drift waves. 
However, the turbulent growth rates are positive for $n=0, m<50$ (subcritical-like behavior), but negative for $n=1$ at all $m$. 
There was already a hint of this from the initial behavior of the curves in Fig.~\ref{transient_n0_n1_growth}, but this is a quantitative and more thorough picture.
Again, the $n=0$ energy injection is a manifestation of the convective filaments drawing energy from the equilibrium density gradient and depositing this energy into the density fluctuations.

@@

\beq
\label{mat_eq}
\mathbf{B}_{m,n} \diff{\mathbf{v}_{m,n}(t)}{t} = \mathbf{C}_{m,n} \mathbf{v}_{m,n}(t) - \sum_{m',n'}  \mathbf{v}_{E,m-m',n-n'} \cdot \gradperp \left( \mathbf{B}_{m',n'} \mathbf{v}_{m',n'}(t) \right),
\eeq
where $\mathbf{v}_{m,n} = \left( N(r_0), N(r_1), \ldots, \vpe(r_0), \vpe(r_1), \ldots, \phi(r_0), \phi(r_1), \ldots, T_e(r_0), T_e(r_1), \ldots \right)_{m,n}^{T}$ is the state of the system,
and $\mathbf{B}_{m,n}$ and $\mathbf{C}_{m,n}$ are coefficient matrices that include the equilibrium information and finite difference coefficients. The first term on the RHS represents the linearities,
and the second term the nonlinearities.
The equations for different $m$ and $n$ are separated, although the nonlinearity couples the different $m$'s and $n$'s. 
In other words, eigenmodes with different $m$ and $n$ are orthogonal and linearly non-interacting. Note that for each $m,n$, there exist $4 \times N_r$ linearly
independent, but nonorthogonal eigenvectors. Hence forth, we drop the $m,n$ subscripts.

@@

We illustrate $G_{\rm{max}}(t)$ along with $G(t)$ for several random initial conditions in Fig.~\ref{m20n0_time_evs} a), all at $n=0$ and $m=20$. Despite the fact that all eigenmodes at this
wavenumber are stable, there is the possibility for transient amplification by a factor of $20$. Further, a sample of randomly initialized vectors are all transiently amplified, but none display
optimal growth as compared to $G_{\rm{max}}(t)$. We show a zoomed-in view of these randomly initialized curves in Fig.~\ref{m20n0_time_evs} b).
It appears extremely unlikely to choose a random initial vector that either optimally amplifies the energy at any time or one that monotonically decays, which would be the case
if the initial vector were one of the linear eigenvectors.

\begin{figure}
\centerline{\includegraphics[]{m20n0_time_evs}}
\caption{a) The maximum growth ratio curve $G_{\rm{max}}(t)$ (dashed line) for $n=0,m=20$, and an ensemble of growth ratio curves (solid gray lines, with the solid black line the ensemble average)
that start with a random initial condition $u(0)$ and evolve under the linear operator. b) A zoomed in version of the ensemble of growth ratio curves. c) A further zoomed in version of the ensemble of
curves with a time range of $\omega_s^{-1}$, which is the characteristic linear time scale for drift waves. 
d) The instantaneous growth rate as a function of time of the the ensemble averaged growth ratio along with the average value over the time spanning $\omega_s^{-1}$.}
\label{m20n0_time_evs}
\end{figure}

@@

Now, since we do not know, \emph{a priori} at what level the turbulence will saturate, we cannot calculate the relative magnitude of $\tau_{\rm{nl}}$ from $(v_E k_\perp)^{-1}$ directly.
Thus, we invoke the conjecture of \emph{critical balance}, which posits that the nonlinear time scale equals the linear time scale at all spatial scales~\cite{schekochihin2012}. 
We can then estimate the nonlinear time scale as $\tau_{\rm{nl}} = \omega_s^{-1}$ -- the inverse of the linear frequency (at each $m,n$).
However, $\omega_s = 0$ for $n=0$ eigenmodes, so we are forced to use $\omega_s$ for the fastest growing $n=1$ eigenmode at each $m$ to get a meaningful time scale. This is not wholly
unjustified, however, due to the inherent nonlinear coupling of different modes, which is evident in Eq.~\ref{mat_eq}.

To illustrate our procedure further, we show in Fig.~\ref{m20n0_time_evs} c) the randomly initialized linear evolution curves from start up until a time of $\omega_s^{-1}$. 
Then, in Fig.~\ref{m20n0_time_evs} d), we calculate the instantaneous growth rate as a function of time of the ensemble averaged curve over that time period.
The average of this instantaneous growth rate curve (the dashed line) gives a ``Transient Growth Rate.''
In Fig.~\ref{spec_nl_transient_gamma}, we compare the Transient Growth Rate Spectra (labeled ``Trans. $\omega_s^{-1}$'') to the Turbulent Growth Rate Spectra.
The $n=0$ curves in Fig.~\ref{spec_nl_transient_gamma} a) are fairly similar, especially when compared to the eigenmode spectra.
The $n=1$ curves in Fig.~\ref{spec_nl_transient_gamma} b), on the other hand, are not very similar, although they trend with $m$ in the same way, which is very different from the linear eigenmode trend.
The difference between the $n=1$ transient and turbulent growth rate spectra is due to nonlinear effects not captured by the simple transient calculation procedure. In particular, the nonlinear
simulation has a strong energy transfer from $n=0$ to $n=1$ Fourier components which is then dissipated by electron-ion collisions as the electrons stream along the field lines~\cite{friedman2012b}.

@@

\begin{figure}
\centerline{\includegraphics[]{spec_nl_transient_gamma}}
\caption{The linear, turbulent, and transient growth rate spectra. The transient growth rate spectra are calculated by the average growth rate over a time $\omega_s^{-1}(m,n=1)$ for the ensemble average
growth ratio curves with the cascade curve using $\omega_s^{-1}$ at $m/2$ rather than $m$.}
\label{spec_nl_transient_gamma}
\end{figure}

@@

\bibliography{refs}
%\bibliographystyle{unsrt}

>>>>

%\documentclass[showpacs,preprintnumbers,amsmath,amssymb,superscriptaddress,aip]{revtex4-1}
\documentclass[letter,scriptaddress,twocolumn, prl,showkeys]{revtex4}
\usepackage{graphicx}
%\setlength{\parindent}{0in}

@@

\title{A Linear Technique to Predict Non-Normal Turbulence Applied to a Magnetized Plasma}

@@

\begin{abstract}
In nonlinear dynamical systems with highly nonorthogonal linear eigenvectors, linear non-modal analysis is more useful than normal mode analysis in predicting turbulent properties. 
However, the non-trivial time evolution of non-modal structures makes quantitative prediction difficult. We present a technique to overcome this difficulty by modelling the effect that the advective nonlinearities have on spatial turbulent structures. The nonlinearities are taken as a periodic randomizing force with
period consistent with critical balance arguments. We apply this technique to a model of drift wave turbulence in the Large Plasma Device (LAPD) 
[W. Gekelman \emph{et al.}, Rev. Sci. Inst. {\bf 62}, 2875 (1991)], where non-modal effects dominate the turbulence.
We compare the resulting growth rate spectrum to that obtained from a nonlinear simulation, showing good quantitative agreement.
\end{abstract}

@@

Normal mode analysis -- the calculation of eigenvalues and eigenvectors of a linearized dynamical system -- has long been used to solve a number of problems ranging from
heat conduction along a solid bar to quantum mechanical energy states of atoms. Despite its wide-ranging success, normal mode analysis has
failed in important instances, particularly in predicting the onset of subcritical turbulence in hydrodynamic flows. 
The reason for this failure was not explained until the early 1990's when Trefethen and others attributed the pitfalls of normal mode analysis to the non-normality of linear operators of
dynamical systems~\cite{trefethen1993,trefethen2005, schmid2007}. A non-normal operator has 
eigenvectors that are nonorthogonal to one another. One consequence of eigenvector nonorthogonality is that even when all eigenvectors are stable and decay exponentially under linear evolution, 
superpositions of eigenvectors can grow, albeit transiently.
Physically, this means that certain types of fluctuations of the laminar state can access free energy from background gradients even though normal mode fluctuations cannot.
When combined with nonlinear effects, this allows for sustained subcritical turbulence.
Such behavior is obscured by traditional normal mode analysis, which only effectively describes the long time asymptotic behavior of fluctuations under the 
action of the linear operator. Transient growth events, which can dominate nonlinear evolution, require initial-value (non-modal) calculations rather than normal mode calculations.

Non-modal analysis has been embraced by the hydrodynamics community over the past two decades in the attempt to explain and predict subcritical turbulence. But the plasma community
still heavily relies on normal mode analysis to inform turbulent predictions and observations, with a few notable exceptions~\cite{camargo1998,camporeale2010,schekochihin2012}. 
Furthermore, non-modal treatments have generally been explanatory rather than predictive and have centered around the transition to turbulence in subcritical systems rather 
than on properties of fully-developed turbulence.
This paper takes up the task of developing a quantitative approach to predicting turbulent properties using only non-modal linear calculations. 
Our approach is to evolve the linear solution from an ensemble of random initial conditions 
and calculate the average growth rate of the solutions over a specified timescale.  Optimally this time scale would be the
nonlinear decorrelation time of the turbulent system, but in order to enable predictive capability, we employ critical balance
arguments to use a characteristic \emph{linear} time.
This procedure produces an effective growth rate spectrum that can be used to predict turbulent properties such as saturation levels.
While the concepts behind this technique are general enough to be applied to many nonlinear dynamical systems, the details vary for each case, 
so we restrict our treatment to one particular turbulence model. The technique reproduces the turbulent growth rate spectrum of the direct nonlinear simulation fairly well, especially
in comparison to the linear spectrum, which has little bearing on the turbulence.

The model describes pressure-gradient-driven turbulence in the uniformly magnetized, cylindrical plasma
produced by the Large Plasma Device (LAPD)~\cite{gekelman1991}. 
Due to LAPD's large size and high collisionality, a fluid model is
appropriate to describe the turbulence~\cite{Popovich2010a,Popovich2010b,Umansky2011}. We use a reduced Braginskii~\cite{Braginskii1965} 2-fluid model:

\beqar
\label{ni_eq}
\pdt N = - {\mathbf v_E} \cdot \grad N_0 - N_0 \gradpar \vpe + \mu_N \gradperp^2 N + S_N + \{\phi,N\}, \\
\label{ve_eq}
\pdt \vpe = - \fmie \frac{T_{e0}}{N_0} \gradpar N - 1.71 \fmie \gradpar T_e  \nonumber \\
+ \fmie \gradpar \phi - \nue \vpe + \{\phi,\vpe \}, \\
\label{rho_eq}
\pdt \varpi = - N_0 \gradpar \vpe  - \nuin \varpi + \mu_\phi \gradperp^2 \varpi + \{\phi,\varpi \}, \\
\label{te_eq}
\pdt T_e = - {\mathbf v_E} \cdot \grad T_{e0} - 1.71 \frac{2}{3} T_{e0} \gradpar \vpe + \frac{2}{3 N_0} \kpe \gradpar^2 T_e  \nonumber \\
- \frac{2 m_e}{m_i} \nue T_e  + \mu_T \gradperp^2 T_e +  S_T + \{\phi,T_e\},
\eeqar
where $N$ is the density, $\vpe$ the parallel electron velocity, $\varpi \equiv \gradperp \cdot (N_0 \gradperp \phi)$ is potential vorticity,
and $T_e$ the electron temperature. The equations employ Bohm normalizations: lengths are
normalized to the ion sound gyroradius $\rho_s$, times to the ion cyclotron time $\omega_{ci}^{-1}$, velocities to the sound speed $c_s$, densities to the equilibrium peak density, and electron
temperatures and potentials to the equilibrium peak electron temperature. The profiles $N_0$ and $T_{e0}$ and other parameters are all taken from experimental measurements. $\phi_0$ 
does not appear in the equations because we model one particular
experiment in which the equilibrium radial electric field and
therefore azimuthal flow was reduced to near zero value using a
limiter biasing technique~\cite{schaffner2012}.

The equations are global and partially linearized. The only nonlinearities retained are the advective nonlinearities, which are written with Poisson brackets. 
We add artificial diffusion and viscosity terms with small numerical coefficients ($\mu_N, \mu_\phi, \mu_T = 10^{-3}$)
to ensure numerical stability in nonlinear simulations, which are performed with the BOUT++ code~\cite{dudson2009}. We use periodic axial and zero value radial 
boundary conditions. Further details of the model, including validation studies, may be found in the references~\cite{Popovich2010a,Popovich2010b,Umansky2011,friedman2012b,friedman2013},
though we mention here that the model does very well in reproducing
the statistical properties of the experimentally-observed turbulence.

The nonlinear simulation reveals a fascinating property of the turbulence -- it is dominated by a nonlinear instability process despite being linearly unstable to drift waves~\cite{friedman2012b,friedman2013}.
This nonlinear instability process, which was discovered by Drake et al.~\cite{drake1995} in a similar plasma model, works as follows: 
magnetic-field-aligned ($k_\para=0$) convective filaments transport density across the equilibrium density gradient, setting up $k_\para=0$ density filaments (the same applies to the temperature). 
These filaments are unstable to a secondary drift wave instability,
which grows on the periphery of the filaments. 
These drift waves, which have finite $k_\para$ and propagate primarily
radially, nonlinearly couple to one another and reinforce the original
convective filaments.

@@

\beq
\label{dEdt_def}
\diff{E(m,n)}{t} = Q(m,n) + D(m,n) + \sum_{m',n'} T(m,m',n,n').
\eeq
In this equation, $m$ and $n$ represent the azimuthal and axial Fourier mode numbers, respectively. 
Since Eqs.~\ref{ni_eq}-\ref{te_eq} are not explicitly dependent on $\theta$ and $z$, the energy equation is separable for every $m$ and $n$. 
Futher, $Q$ represents the linear energy exchange with the equilibrium (convective transport), $D$ the dissipation from irreversible processes
like collisions, and $T$ the nonlinear triad wave transfer. Since $\int f \{g,f\} dV = 0$, or equivalently $\sum_{m,m',n,n'} T(m,m',n,n')=0$, 
the Poisson bracketed advective nonlinearities in the model are energetically conservative and therefore do no inject or dissipate energy on the whole.
Consequently, the fluctuation energy originates completely from $Q$, the linear convective transport mechanism.

\begin{figure}
\centerline{\includegraphics[width=0.4\textwidth]{transient_n0_n1_growth}}
\caption{Linear evolution of energy starting from a turbulent initial state. The $n=0$ curves have an initial period of transient growth before exponentially decaying.}
\label{transient_n0_n1_growth}
\end{figure}

This convective transport of density filaments is akin to the paradigmatic ``lift-up'' mechanism in hydrodynamic shear flows whereby streamwise vortices drive streamwise streaks~\cite{krommes1999}.
Both are transient growth processes. We see this in our simulations by following the evolution of the energy of several Fourier $m,n$ modes after turning off the nonlinearities in an already
turbulent simulation. A few representative $m,n$ modes are shown in
Fig.~\ref{transient_n0_n1_growth}. The linear transient growth of the filamentary structures is seen in the $n=0$ modes, 
which grow transiently before decaying exponentially at the rate of their least stable eigenmode. This kind of behavior is rather unituitive to some because all $n=0$ linear eigenmodes are stable, 
and the normal mode paradigm has conditioned us to believe that a linear superposition of such eigenmodes should decay. But as the hydrodynamics community discovered, linear
superpositions of nonorthogonal vectors may behave unintuitively.
Notice now that the $n=1,m=20$ mode decays transiently before growing exponentially with growth rate of the most unstable eigenmode at $n=1,m=20$. This isn't quite as interesting as the $n=0$
behavior, and it can occur in normal systems. Futhermore, it has no parallel in truly subcritical turbulence, but it can be helpful for interpreting later results.

@@

\begin{figure}
\centerline{\includegraphics[width=0.45\textwidth]{spec_nl_gamma}}
\caption{Linear and turbulent growth rate $m$ spectra for $n=0$ (solid lines) and $n=1$ (dashed lines) Fourier components. The linear growth rates are those of the least stable eigenmodes at each
$m,n$, while the turbulent growth rates represent $\pdiff{E}{t}|_{\rm{lin}}/2 E$ from the nonlinear simulation. The shaded region marks the $1 \sigma$ spread in the turbulent growth rate spectrum,
obtained from the distribution of growth rates calculated from the nonlinear simulation over a long time range.}
\label{spec_nl_gamma}
\end{figure}

To obtain the turbulent growth rate spectrum from the nonlinear simulation, we begin with the expression $\diff{E(m,n)}{t}\big|_{\rm{lin}} = Q(m,n) + D(m,n)$, 
which accounts for the total net energy injection at each $m,n$ into the fluctuations. 
An effective instantaneous growth rate $\gamma_{\rm{eff}}(m,n,t) = \diff{E(m,n,t)}{t}\big|_{\rm{lin}}/ 2 E(m,n,t)$ can then be defined. 
This effective growth rate is calculated from the spatial structure of the fluctuations, so it is well-defined for both turbulent fluctuations and eigenmode fluctuations. If calculated
from a linear simulation run for a sufficiently long time, $\gamma_{\rm{eff}}(m,n)$ is the same as the eigenmode growth rate of the least stable linear eigenmode for each $m,n$.
For turbulence, $\gamma_{\rm{eff}}(m,n)$ is generally not equal to an eigenmode growth rate, especially in the case where the system is non-normal.
We plot this growth rate in Fig.~\ref{spec_nl_gamma} for $n=0$ and $n=1$ components. The turbulent growth rate curves are calculated from this formula, 
averaged over a time of about $3 \times 10^{4} \ \omega_{ci}^{-1}$ during the saturated turbulent phase of the nonlinear simulation: $\gamma_{\rm{Turb}}(m,n) = \int \gamma_{\rm{eff}}(m,n,t) dt$. 
We show this turbulent growth rate along with the $1 \sigma$ contour, which indicates how greatly $\gamma_{\rm{eff}}(m,n,t)$ varies over time.
Along with these, the growth rates of the least stable (or most unstable) linear eigenmodes are shown.

As already stated, one sees from Fig.~\ref{spec_nl_gamma} that the linear eigenmodes are unstable for $n=1$ and stable for $n=0$ as expected for drift waves. 
However, the turbulent growth rates are positive for $n=0, m<50$ (subcritical-like behavior), but negative for $n=1$ at all $m$. 
There was already a hint of this from the initial behavior of the
curves in Fig.~\ref{transient_n0_n1_growth}, but this figure represents
quantitative confirmation.
Again, the $n=0$ energy injection is a manifestation of the convective filaments drawing energy from the equilibrium density gradient and depositing this energy into the density fluctuations.

@@

\beqar
\label{mat_eq}
\mathbf{B}_{m,n} \diff{\mathbf{v}_{m,n}(t)}{t} = \mathbf{C}_{m,n} \mathbf{v}_{m,n}(t) \nonumber \\
- \sum_{m',n'}  \mathbf{v}_{E,m-m',n-n'} \cdot \gradperp \left( \mathbf{B}_{m',n'} \mathbf{v}_{m',n'}(t) \right),
\eeqar
where $\mathbf{v}_{m,n} = \left( N(r), \vpe(r), \phi(r), T_e(r) \right)_{m,n}^{T}$ is the state of the system,
and $\mathbf{B}_{m,n}$ and $\mathbf{C}_{m,n}$ are coefficient matrices that include the equilibrium information and finite difference coefficients. The first term on the RHS represents the linearities,
and the second term the nonlinearities.
The equations for different $m$ and $n$ are separated, although the nonlinearity couples the different $m$'s and $n$'s. 
In other words, eigenmodes with different $m$ and $n$ are orthogonal and linearly non-interacting. Note that for each $m,n$, there exist $4 \times N_r$ linearly
independent, but nonorthogonal eigenvectors. Hence forth, we drop the $m,n$ subscripts.

@@

We illustrate $G_{\rm{max}}(t)$ along with $G(t)$ for several random initial conditions in Fig.~\ref{m20n0_time_evs} a), all at $n=0$ and $m=20$. Despite the fact that all eigenmodes at this
wavenumber are stable, there is the possibility for transient amplification by a factor of $20$. Furthermore, a sample of randomly initialized vectors are all transiently amplified, but none display
optimal growth as compared to $G_{\rm{max}}(t)$. We show a zoomed-in view of these randomly initialized curves in Fig.~\ref{m20n0_time_evs} b).
It appears extremely unlikely to choose a random initial vector that either optimally amplifies the energy at any time or one that monotonically decays, which would be the case
if the initial vector were one of the linear eigenvectors.

\begin{figure}
\centerline{\includegraphics[width=0.5\textwidth]{m20n0_time_evs}}
\caption{a) The maximum growth ratio curve $G_{\rm{max}}(t)$ (dashed line) for $n=0,m=20$, and an ensemble of growth ratio curves (solid gray lines, with the solid black line the ensemble average)
that start with a random initial condition $u(0)$ and evolve under the linear operator. b) The ensemble of growth ratio curves alone, and c) cut off at $t=\omega_s^{-1}$, the drift wave eigenmode
time scale. d) The instantaneous growth rate of the ensemble averaged growth ratio along with the average value.}
\label{m20n0_time_evs}
\end{figure}

@@

Now, since we do not know \emph{a priori} at what level the turbulence will saturate, we cannot calculate the relative magnitude of $\tau_{\rm{nl}}$ from $(v_E k_\perp)^{-1}$ directly.
Thus, we invoke the conjecture of \emph{critical balance}, which posits that the nonlinear time scale equals the linear time scale at all spatial scales~\cite{schekochihin2012}. 
We can then estimate the nonlinear time scale as $\tau_{\rm{nl}} = \omega_s^{-1}$ -- the inverse of the linear frequency (at each $m,n$).
However, $\omega_s = 0$ for $n=0$ eigenmodes, so we are forced to use $\omega_s$ for the fastest growing $n=1$ eigenmode at each $m$ to get a meaningful time scale. This is not wholly
unjustified, however, due to the inherent nonlinear coupling of different modes, which is evident in Eq.~\ref{mat_eq}.

To illustrate our procedure further, we show in Fig.~\ref{m20n0_time_evs} c) the randomly initialized linear evolution curves from start up until a time of $\omega_s^{-1}$. 
The transient growth is cut off well before optimal amplification occurs, but as reflected by our procedure, there is no guarantee that the nonlinearities will allow complete non-modal amplification.
Then, in Fig.~\ref{m20n0_time_evs} d), we calculate the instantaneous growth rate as a function of time of the ensemble averaged curve over that time period.
The average of this instantaneous growth rate curve (the dashed line) gives a ``Transient Growth Rate.''
In Fig.~\ref{spec_nl_transient_gamma}, we compare the Transient Growth Rate Spectra (labeled ``Trans. $\omega_s^{-1}$'') to the Turbulent Growth Rate Spectra.
The $n=0$ curves in Fig.~\ref{spec_nl_transient_gamma} a) are fairly similar, especially when compared to the eigenmode spectra.
The $n=1$ curves in Fig.~\ref{spec_nl_transient_gamma} b), on the other hand, are not very similar, although they trend with $m$ in the same way, which is very different from the linear eigenmode trend.
The difference between the $n=1$ transient and turbulent growth rate spectra is due to nonlinear effects not captured by the simple transient calculation procedure. In particular, the nonlinear
simulation has a strong energy transfer from $n=0$ to $n=1$ Fourier components which is then dissipated by electron-ion collisions as the electrons stream along the field lines~\cite{friedman2012b}.

@@

\begin{figure}
\centerline{\includegraphics[width=0.5\textwidth]{spec_nl_transient_gamma}}
\caption{The linear, turbulent, and transient growth rate spectra. The transient growth rate spectra are calculated by the average growth rate over a time $\omega_s^{-1}(m,n=1)$ for the ensemble average
growth ratio curves with the cascade curve using $\omega_s^{-1}$ at $m/2$ rather than $m$.}
\label{spec_nl_transient_gamma}
\end{figure}

@@

%\bibliographystyle{unsrt}
\bibliography{refs}
